---
output: github_document
---

# BayesRTMB

**Language / Ť®ÄŤ™ě**: [English introduction](articles/introduction.html) | [„ÉĎ„ÉÉ„āĪ„Éľ„āłÁīĻšĽč](articles/ja-introduction.html)

**BayesRTMB** is an R package for writing and fitting statistical models with RTMB as the automatic differentiation engine.

You can start from wrapper functions such as `rtmb_lm()`, `rtmb_glmer()`, `rtmb_corr()`, and `rtmb_ttest()`, or write your own model with `rtmb_code()`. The same model object can then be used for MCMC, MAP estimation, variational inference, and frequency-oriented classical analyses where supported.

## Key Features

- **Model code in R**: Write models with `setup`, `parameters`, `transform`, `model`, and `generate` blocks.
- **Multiple estimation methods**: Use `$sample()`, `$optimize()`, `$variational()`, and `$classic()` from a common model object.
- **Wrapper functions**: Fit regression models, GLM/GLMMs, t tests, correlations, contingency tables, factor analysis, IRT, latent rank models, and multidimensional unfolding models.
- **Classical analyses from wrappers**: Use `$classic()` for frequency-oriented outputs, including `AIC()`, `BIC()`, and `anova()` for supported fits.
- **Random effects and Laplace approximation**: Declare parameters with `random = TRUE` and use RTMB's Laplace machinery for latent variables and mixed models.
- **Diagnostics and visualization**: Plot posterior draws, convergence diagnostics, forest plots, conditional effects, and related summaries.

## Installation

You can install the development version from GitHub with either `pak` or `remotes`.

```{r eval=FALSE}
pak::pak("norimune/BayesRTMB")
```

```{r eval=FALSE}
remotes::install_github("norimune/BayesRTMB")
```

### Windows Users

Windows users need Rtools because BayesRTMB depends on the RTMB/TMB ecosystem.

```{r eval=FALSE}
pkgbuild::check_build_tools(debug = TRUE)
```

If this check fails, install the Rtools version that matches your R version from the [Rtools page](https://cran.r-project.org/bin/windows/Rtools/), restart R, and try again.

## Quick Example

For standard analyses, start with a wrapper function.

```{r eval=FALSE}
library(BayesRTMB)
data(debate)

mdl <- rtmb_lm(sat ~ talk * perf, data = debate)

fit_mcmc <- mdl$sample()
fit_map  <- mdl$optimize()
fit_lm   <- mdl$classic()
```

The generated model code can be inspected.

```{r eval=FALSE}
mdl$print_code()
```

You can also write a model directly.

```{r eval=FALSE}
data_list <- list(Y = debate$sat, N = nrow(debate))

code <- rtmb_code(
  parameters = {
    mu = Dim()
    sigma = Dim(lower = 0)
  },
  model = {
    Y ~ normal(mu, sigma)
    mu ~ normal(0, 10)
    sigma ~ exponential(1)
  }
)

mdl_custom <- rtmb_model(data_list, code)
fit_custom <- mdl_custom$sample()
```

## Articles

- [Quick Start](articles/quick_start.html): installation, a minimal model, MCMC diagnostics, visualization, and t tests.
- [Wrapper Functions](articles/wrapper_functions.html): using wrappers for Bayesian and classical analyses.
- [Writing Model Codes](articles/writing_models.html): writing custom models with `rtmb_code()`.
- [śó•śú¨Ť™ě: „ÉĎ„ÉÉ„āĪ„Éľ„āłÁīĻšĽč](articles/ja-introduction.html)
- [śó•śú¨Ť™ě: „āĮ„ā§„ÉÉ„āĮ„āĻ„āŅ„Éľ„Éą](articles/ja-quick_start.html)
- [śó•śú¨Ť™ě: „É©„ÉÉ„ÉĎ„ÉľťĖĘśēį„ĀģšĹŅ„ĀĄśĖĻ](articles/ja-wrapper_functions.html)
- [śó•śú¨Ť™ě: „ÉĘ„Éá„Éę„ā≥„Éľ„ÉČ„Āģśõł„ĀćśĖĻ](articles/ja-writing_models.html)
