---
title: "Introduction to BayesRTMB"
description: "An overview of the concepts, basic usage, and main inference methods in BayesRTMB."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BayesRTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# What is BayesRTMB?

**BayesRTMB** is an R package for Bayesian inference that leverages the RTMB automatic differentiation engine. It allows you to write models using a Stan-like intuitive syntax while performing inference entirely within R, eliminating the need for C++ compilation.

This page provides an overview of BayesRTMB and the basic workflow you should know to get started.

## Key Features

BayesRTMB offers the following advantages:

- **No C++ compilation**: Models defined with `rtmb_code()` are evaluated instantly in R.
- **Organized block syntax**: Write your model using structured blocks (`setup`, `parameters`, `transform`, `model`, etc.).
- **Unified interface**: Easily switch between NUTS (MCMC), ADVI (Variational Inference), and MAP estimation from a single model object.
- **Laplace approximation**: Efficiently marginalize random effects in hierarchical models.
- **Model evaluation**: Built-in support for computing marginal likelihoods and Bayes factors via Bridge Sampling.
- **Convenient wrapper functions**: Start analyzing data immediately with functions like `rtmb_lm()` and `rtmb_glmer()` without writing custom models from scratch.

## Installation

You can install the development version from GitHub using the `pak` package:

```{r eval=FALSE}
# install.packages("pak")
pak::pak("norimune/BayesRTMB")
library(BayesRTMB)
```

## Basic Workflow

A typical analysis in BayesRTMB follows these 3 steps:

1. **Prepare the data** as a list.
2. **Define the model** using `rtmb_code()`.
3. **Create the model object** with `rtmb_model()` and perform estimation.

Let's look at a simple example of estimating the mean `mu` and standard deviation `sigma` of a normal distribution.

```{r eval=FALSE}
# 1. Prepare data
Y <- c(5.2, 4.8, 5.5, 6.1, 4.9, 5.3)
dat <- list(Y = Y)

# 2. Define the model
code <- rtmb_code(
  parameters = {
    mu    = Dim(1)
    sigma = Dim(1, lower = 0)
  },
  model = {
    # Priors
    mu    ~ normal(0, 10)
    sigma ~ exponential(0.1)
    # Likelihood
    Y     ~ normal(mu, sigma)
  }
)

# 3. Create the model object
mdl <- rtmb_model(data = dat, code = code)
```

## Choosing an Estimation Method

Once you have created the `mdl` object, you can call multiple estimation methods depending on your needs.

### 1. MAP Estimation (`optimize`)
Finds the Maximum A Posteriori (MAP) estimate. Because it is extremely fast, it is highly recommended for initially checking your model for syntax errors or verifying the general behavior.

```{r eval=FALSE}
fit_map <- mdl$optimize()
fit_map$summary()
```

### 2. MCMC Estimation (`sample`)
Performs MCMC sampling using the NUTS (No-U-Turn Sampler) algorithm to obtain exact samples from the posterior distribution. This is the standard approach for rigorous Bayesian inference. Parallel computation across multiple chains is supported via `parallel = TRUE`.

```{r eval=FALSE}
fit_mcmc <- mdl$sample(sampling = 1000, warmup = 1000, chains = 4, parallel = FALSE)
fit_mcmc$summary()
```

### 3. Variational Inference (`variational`)
Approximates the posterior distribution using ADVI (Automatic Differentiation Variational Inference). It is useful when you want to quickly estimate complex models where MCMC would take too much time. However, be aware that uncertainty (standard errors) might be underestimated.

```{r eval=FALSE}
fit_vb <- mdl$variational(
  method = "meanfield",
  iter = 3000,
  num_estimate = 4
)
fit_vb$summary()
```

## Models with Random Effects

When building hierarchical models, you can declare parameters as random effects using `random = TRUE`.

```{r eval=FALSE}
code_hier <- rtmb_code(
  parameters = {
    mu_global    = Dim(1)
    sigma_global = Dim(1, lower = 0)
    alpha        = Dim(J, random = TRUE)
  },
  model = {
    alpha        ~ normal(mu_global, sigma_global)
    # ... likelihood ...
  }
)
```

For such models, the **Laplace approximation** can be used to marginalize (integrate out) the random effects. 
This is enabled by default in `optimize(laplace = TRUE)`. While it can also be used in MCMC, `laplace = FALSE` is usually preferred for full sampling.

```{r eval=FALSE}
fit_map_laplace  <- mdl$optimize(laplace = TRUE)
fit_mcmc_laplace <- mdl$sample(sampling = 1000, warmup = 1000, chains = 4, laplace = FALSE)
```

## Quick Start with Wrapper Functions

BayesRTMB provides several wrapper functions to quickly run standard analyses without defining the model from scratch.

For example, to run a linear regression, you can use `rtmb_lm()`:

```{r eval=FALSE}
fit_lm <- rtmb_lm(mpg ~ wt + cyl, data = mtcars)

# MAP Estimation
map_lm <- fit_lm$optimize()
map_lm$summary()

# MCMC Estimation
mcmc_lm <- fit_lm$sample(sampling = 1000, warmup = 1000, chains = 4)
mcmc_lm$summary()
```

Other functions like `rtmb_glmer()` for mixed models and `rtmb_fa()` for factor analysis are also available.

You can also perform a Bayesian t-test effortlessly using `rtmb_ttest()`. By specifying the parameter you want to fix to 0 in the `null_model` argument of the `bayes_factor()` method, the null model is automatically estimated and compared.

```{r eval=FALSE}
mdl_ttest <- rtmb_ttest(y1, y2, r = 0.707)
mcmc_ttest <- mdl_ttest$sample()

# Calculate Bayes Factor against the null model (effect size delta = 0)
bf_ttest <- mcmc_ttest$bayes_factor(null_model = "delta")
print(bf_ttest)
```

## Next Steps

Once you grasp the overview of BayesRTMB, we recommend checking the following documentation for deeper understanding:

1. **Quick Start (ja-quick_start.html / Available in Japanese)**
   Learn the roles of each block (`setup`, `transform`, etc.) and practical workflows through examples of binomial, regression, and hierarchical models.
2. **Reference**
   Check detailed specifications for each function. Especially when building custom models, the following pages are helpful:
   - `rtmb_code()`: Block syntax rules
   - `Dim()`: Parameter types and constraints (`parameter_types`)
   - `distributions` / `math_functions`: Built-in probability distributions and stable mathematical utilities
